Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-01-13 2.background/2.RV_and_RS.tex

In the case of independent and identically distributed $\epsilon_p(i)$'s, \citet{barndorff1963limit} and \citet{resnick1973almost} obtained necessary and sufficient conditions for the \emph{almost sure stability} of maxima, where the convergence in \eqref{eq:RS-condition} holds almost surely. \stilian{See also \cite{klass:1984} for further 
sharp results on almost sure stability.  Here, we will only need the weaker notion in \eqref{eq:RS-condition} 
but extend our inquiry to the case of dependent $\epsilon_p(i)$'s.}{\fbox{The ref suggested by a reviewer -- Resnick?}}

@@

\begin{proposition}[Rapid variation and relative stability] \label{prop:rapid-varying-tails}
Assume that the array ${\cal E}$ consists of identically distributed and possibly dependent random 
variables with cumulative distribution function $F$, where $F(x)<1$ for all finite $x>0$. 
\begin{enumerate}
    \item If $F$ has rapidly varying right tail in the sense of \eqref{e:def:rapid-variation}, then for all $\delta>0$,
        \begin{equation} \label{eq:rapid-varying-tails}
            \P\left[\frac{1}{u_p} M_p\le1+\delta\right] \ge 1 - \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)} \to 1.
        \end{equation}
    \item If the array ${\cal E}$ has independent entries, then it is relatively stable if and only if $F$ has rapidly varying tail, i.e., \eqref{def:rapid-variation} holds.
    \label{prop:rapid-varying-tails_part-ii}
\end{enumerate}
\end{proposition}

@@

\stilian{ 
\begin{remark} \label{rem:iid-max} Part (1) of Proposition \ref{prop:rapid-varying-tails} is equivalent to 
\begin{equation}\label{e:rem:iid-max}
  \P\left[\frac{1}{u_p} M_p >  1+\delta_p \right] \longrightarrow 0,\ \ \mbox{ as }p\to\infty,
\end{equation}
for some positive sequence $\delta_p\to 0$.  Notice on the other hand that if $M_p^*$ is the maximum of $p$ iid variables with distribution $F$,
the relative stability property entails $M_p^*/u_p \to 1$, in probability, as $p\to\infty$.  Since the sequence $1+\delta_p\to 1$, 
Relation \eqref{e:rem:iid-max} means that the rate of growth of the maxima $M_n$ in ${\cal E}$ cannot be faster than that of the independent
maxima $M_p^*$.  This somewhat surprising fact is true regardless of the dependence structure of ${\cal E}$ and is solely a 
consequence of the rapid variation of $F$.
\end{remark}
}{\fbox{The above remark addresses a question of a reviewer}}

>>>>

In the case of independent and identically distributed $\epsilon_p(i)$'s, \citet{barndorff1963limit} and \citet{resnick1973almost} obtained necessary and sufficient conditions for the \emph{almost sure stability} of maxima, where the convergence in \eqref{eq:RS-condition} holds almost surely. See also \cite{klass:1984} for further sharp results on almost sure stability, and \cite{naveau2003almost} for almost sure stability in stationary sequences. 
Here, we will only need the weaker notion in \eqref{eq:RS-condition} but extend our inquiry to the case of dependent $\epsilon_p(i)$'s.

@@

\begin{proposition}[Rapid variation and relative stability] \label{prop:rapid-varying-tails}
Assume that the array ${\cal E}$ consists of identically distributed and possibly dependent random 
variables with cumulative distribution function $F$, where $F(x)<1$ for all finite $x>0$. 
\begin{enumerate}
    \item If $F$ has rapidly varying right tail in the sense of \eqref{e:def:rapid-variation}, then for all $\delta>0$,
        \begin{equation} \label{eq:rapid-varying-tails}
            \P\left[\frac{1}{u_p} M_p\le1+\delta\right] \ge 1 - \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)} \to 1.
        \end{equation}
    \item If the array ${\cal E}$ has independent entries, then it is relatively stable if and only if $F$ has rapidly varying tail, i.e., \eqref{e:def:rapid-variation} holds.
    \label{prop:rapid-varying-tails_part-ii}
\end{enumerate}
\end{proposition}

@@

\begin{remark} \label{rem:iid-max} Part (1) of Proposition \ref{prop:rapid-varying-tails} is equivalent to 
\begin{equation}\label{e:rem:iid-max}
  \P\left[\frac{1}{u_p} M_p >  1+\delta_p \right] \longrightarrow 0,\ \ \mbox{ as }p\to\infty,
\end{equation}
for some positive sequence $\delta_p\to 0$.  
Notice on the other hand that, if $M_p^*$ is the maximum of $p$ iid variables with distribution $F$,
the relative stability property entails $M_p^*/u_p \to 1$, in probability, as $p\to\infty$.  Since the sequence $1+\delta_p\to 1$, 
Relation \eqref{e:rem:iid-max} means that the rate of growth of the maxima $M_n$ in ${\cal E}$ cannot be faster than that of the independent
maxima $M_p^*$. 
This somewhat surprising fact holds regardless of the dependence structure of ${\cal E}$ and is solely a consequence of the rapid variation of $F$.
\end{remark}

### 2021-01-13 2.background/2.asymptotics.tex

\medskip
{\bf Signal detection.}
\stilian{The asymptotic behavior of the statistical risk in signal detection problems \eqref{eq:risk-detection} in high dimensions was first studied by Yurii Ismailovich Ingster in the context of sparse additive models \eqref{eq:model-additive} with 
independent and Gaussian components. Specifically, \cite{ingster1998minimax} considered the behavior of the 
theoretically optimal likelihood ratio (\ac{LR}) test in the high-dimensional asymptotic regime, where the dimension $p$ grows to infinity. Then, under certain parameterization of the size and sparsity of the signal $\mu$, there are only two possibilities: Either ${\rm risk}^D(LR)$ vanishes, as $p\to\infty$, or $\liminf_{p\to\infty} {\rm risk}^D(LR) =1$.
The precise signal size and sparsity parameterizations as well as the so-called {\em detection boundary} discovered 
by Ingster are described in Chapter \ref{chap:phase-transitions}.}{\fbox{replacing?}
 where a phase transition phenomenon was discovered for 
That is, as $p\to\infty$, depending on the size and sparsity of the signal $\mu$, the detection risk either vanishes, 
or has a liminf of 1 when we apply the theoretically optimal likelihood ratio (\ac{LR}) test. }

The \ac{LR} test, unfortunately, depends  relies on the knowledge of the signal sparsity and signal sizes which are 
not available in practice. The sparsity-and-signal-size-agnostic statistic \ac{HC} in \eqref{eq:HC-statistic} was identified 
to attain such optimal performance limits in sparse Gaussian models in \cite{donoho2004higher}. 
A modified goodness-of-fit test statistic in \citet{zhang2002powerful}, and two statistics based on thresholded-$L_1$ and $L_2$ norms in \cite{zhong2013tests} were also shown to be asymptotically optimal in the detection problem.
Recent studies have also focused on the behavior of detection risk \eqref{eq:risk-detection} in dense and scale mixture models \cite{cai2011optimal}, under general distributional assumptions \citep{cai2014optimal, arias2017distribution1}, as well as when the errors are dependent \citep{hall2010innovated}.
A comprehensive review focusing on the role of \ac{HC} in detection problems can be found in \cite{donoho2015special}.
Notwithstanding the extensive literature on the detection problem, the performances of simple statistics such as 
$L_q$ norms \eqref{eq:Lq-norm} and sums \eqref{eq:sum-statistic}, to the best of our knowledge, have only been 
sparingly documented.  We gather relevant results in Chapter \ref{chap:phase-transitions}, and make several new 
contributions on the performance of several statistics commonly used in practice.

@@

{\bf Exact support recovery.}
There is a wealth of literature on the so-called sparsistency (i.e., $\P[\widehat{S} = S]\to 1$ as $p\to\infty$) problem in the regression context. 
Sparsistency problems were pursued, among many others, by \citet{zhao2006model} and \citet{wasserman2009high} in the high-dimensional regression setting (where the number of samples $n\ll p$), and by \citet*{meinshausen2006high} in graphical models.
Although there have been numerous studies on the sufficient conditions for sparsistency, efforts on necessary conditions have been scarce.
Notable exceptions include \cite{wainwright2009information, wainwright2009sharp} and \cite{comminges2012tight} in regression problems.
We refer the reader to the recent book by \cite{wainwright2019high} (and in particular, the bibliographical sections of Chapters 7 and 15, therein) for a comprehensive review.

@@

Nevertheless, two important questions remained unanswered.
Namely, precise phase-transition-type results for the exact support recovery risk 
\eqref{eq:risk-exact} and for the support recovery probability \eqref{eq:risk-prob} 
% --- akin to that obtained in \cite{arias2017distribution} for the approximate support recovery risk \eqref{eq:risk-approximate} --- 
have not been established.
And secondly, performance of commonly used statistical procedures reviewed in Section \ref{sec:statistical-procedures}
in terms of these risk metrics have not been studied.
\stilian{Some of our main contributions in this text are the solutions to these problems, presented in Chapters \ref{chap:phase-transitions} and \ref{chap:exact-support-recovery}, below. Specifically, it turns out that the simple Bonferoni thresholding procedure (among others) is asymptotically optimal for the exact support recovery problem in 
\eqref{eq:model-additive} under broad classes of error distributions. Furthermore, a phase-transition in the exact support recovery 
problem for thresholding procedures is established under broad dependence conditions on the 
errors using the concentration of maxima phenomenon (Chapter \ref{chap:exact-support-recovery}). While the 
optimality of thresholding procedures under dependence remains an open question, in Chapter \ref{sec:optimality}, 
we establish finite-sample Bayes optimality and sub-optimality results for these procedures under independence.  
The landscape of the fundamental statistical limits in support estimation is yet to be fully charted. We conjecture, 
however, that the general concentration of maxima phenomenon will lead to its complete solution under very broad error 
dependence scenarios.  }{
Resolving these two issues, we show in this text that several well-known FWER-controlling procedures --- including Bonferroni's procedure --- are optimal in the additive error model under both one-sided and two-sided alternatives.}
% these results have appeared in \cite{gao2018fundamental} and \cite{gao2019five}

@@

In the additive error model \eqref{eq:model-additive} under independent Gaussian errors and one-sided alternatives \eqref{eq:global-test-one-sided},
\cite{arias2017distribution} showed that a phase transition exists for  the approximate support recovery risk \eqref{eq:risk-approximate}.
The \ac{BH} procedure \cite{benjamini1995controlling}, and the \stilian{DISCUSS}{Is it Candes-Barber or Barber-Candes?}
Cand\`es-Barber procedure \citep{barber2015controlling} was identified to be asymptotically optimal. % \citet{rabinovich2017optimal} further established the rate-optimality of both procedures under the same regime.
However, \cite{arias2017distribution}, as all related work so far, assumed the non-nulls to follow a common alternative distribution.
We derive a new phase transition result that relaxes this assumption on the alternatives in Chapter \ref{chap:phase-transitions}.

@@

{\bf Asymmetric statistical risks.}
Although weighted sums of false discovery and non-discovery have been studied in the literature mentioned above, \stilian{the case of simultaneous family-wise error control and marginal, location-wise power requirements, has not been previously considered. }{\fbox{Altered, check the wording.}} As a result, asymmetric statistical risks such as \eqref{eq:risk-exact-approx} and \eqref{eq:risk-approx-exact} have not 
been investigated.
As argued in Section \ref{sec:risks}, the properties of these asymmetric risks are of important practical concern in applications such as GWAS.
We study the asymptotic behavior of these risks in Chapters \ref{chap:phase-transitions} and \ref{chap:GWAS} of this text.
% ; the results therein have appeared in \cite{gao2019five}.
% these results have appeared in \cite{gao2018fundamental} and \cite{gao2019five}

{\bf Chi-square models and GWAS.}
The high-dimensional chi-square model \eqref{eq:model-chisq} seemed to have received little attention in the literature.
While the sparse signal detection problem in the chi-square model has been studied \cite{donoho2004higher}, to the best of our knowledge,  asymptotic limits of the support recovery problems have not been studied.
The chi-squared model and the motivating GWAS application are analyzed in Chapter \ref{chap:GWAS}.
\stilian{The results obtained therein help us explain the sharp power decay in GWAS known as the 
{\em steep part of the power curve}.  This empirical fact has long been observed by statistical geneticists \citep{bush2012genome} but so far it had not been mathematically quantified.  \cite{gao2019upass} provide 
further details on the power and design in GWAS as well as an accompanying interactive 
statistical software \citep{gaoUPASS_ShinyApp}.}{\fbox{altered the text, pls check/edit.}}

>>>>

\medskip
{\bf Signal detection.}
The asymptotic behavior of the statistical risk in signal detection problems \eqref{eq:risk-detection} in high dimensions was first studied by \stilian{Yurii Ismailovich Ingster}{where did you find the full name?} in the context of sparse additive models \eqref{eq:model-additive} with 
independent and Gaussian components. Specifically, \cite{ingster1998minimax} considered the behavior of the 
theoretically optimal likelihood ratio (\ac{LR}) test in the high-dimensional asymptotic regime, where the dimension $p$ grows to infinity. Then, under certain parameterization of the size and sparsity of the signal $\mu$, a dichotomy exists: 
either ${\rm risk}^D(\mathcal{R})$ vanishes as $p\to\infty$ where $\mathcal{R}$ is the \ac{LR} test, or $\liminf_{p\to\infty} {\rm risk}^D(\mathcal{R}) = 1$ for any procedure.
The precise signal size and sparsity parameterizations as well as the so-called {\em detection boundary} discovered by Ingster are described in Chapter \ref{chap:phase-transitions}.

The \ac{LR} test, unfortunately, relies on the knowledge of the signal sparsity and signal sizes which are 
not available in practice. The sparsity-and-signal-size-agnostic statistic \ac{HC} in \eqref{eq:HC-statistic} was identified 
to attain such optimal performance limits in sparse Gaussian models in \cite{donoho2004higher}. 
A modified goodness-of-fit test statistic in \citet{zhang2002powerful}, and two statistics based on thresholded-$L_1$ and $L_2$ norms in \cite{zhong2013tests} were also shown to be asymptotically optimal in the detection problem.
Recent studies have also focused on the behavior of detection risk \eqref{eq:risk-detection} in dense and scale mixture models \cite{cai2011optimal}, under general distributional assumptions \citep{cai2014optimal, arias2017distribution1}, as well as when the errors are dependent \citep{hall2010innovated}.
A comprehensive review focusing on the role of \ac{HC} in detection problems can be found in \cite{donoho2015special}.
Notwithstanding the extensive literature on the detection problem, the performances of simple statistics such as 
$L_q$ norms \eqref{eq:Lq-norm} and sums \eqref{eq:sum-statistic}, to the best of our knowledge, have only been 
sparingly documented.  We gather relevant results in Chapter \ref{chap:phase-transitions}, and make several new 
contributions on the performance of several statistics commonly used in practice.

@@

{\bf Exact support recovery.}
There is a wealth of literature on the so-called sparsistency (i.e., $\P[\widehat{S} = S]\to 1$ as $p\to\infty$) problem in the regression context. 
Sparsistency problems were pursued, among many others, by \citet{zhao2006model} and \citet{wasserman2009high} in the high-dimensional regression setting (where the number of samples $n\ll p$), and by \citet*{meinshausen2006high} in graphical models.
Although there have been numerous studies on the sufficient conditions for sparsistency, efforts on necessary conditions have been scarce.
Notable exceptions include \cite{wainwright2009information, wainwright2009sharp} and \cite{comminges2012tight} in regression problems.
We refer the reader to the recent book by \cite{wainwright2019high} (and in particular, the bibliographical sections of Chapters 7 and 15 therein) for a comprehensive review.

@@

Nevertheless, two important questions remained unanswered.
Namely, precise phase-transition-type results for the exact support recovery risk 
\eqref{eq:risk-exact} and for the support recovery probability \eqref{eq:risk-prob} 
% --- akin to that obtained in \cite{arias2017distribution} for the approximate support recovery risk \eqref{eq:risk-approximate} --- 
have not been established.
And secondly, performance of commonly used statistical procedures reviewed in Section \ref{sec:statistical-procedures}
in terms of these risk metrics have not been studied.
Some of our main contributions in this text are the solutions to these problems, presented in Chapters \ref{chap:phase-transitions} and \ref{chap:exact-support-recovery} below. 
Specifically, we show that the Bonferroni thresholding procedure (among others) is asymptotically optimal for the exact support recovery problem in \eqref{eq:model-additive} under broad classes of error distributions. 
Furthermore, a phase-transition in the exact support recovery problem for thresholding procedures is established under broad dependence conditions on the errors using the concentration of maxima phenomenon (Chapter \ref{chap:exact-support-recovery}). 
We also establish finite-sample Bayes optimality and sub-optimality results for these procedures under independence, and by extension, arrive at minimax optimality results for the exact support recovery problem.

The landscape of the fundamental statistical limits in support estimation is yet to be fully charted. 
We conjecture, however, that the general concentration of maxima phenomenon will lead to its complete solution under very broad error 
dependence scenarios.
\medskip

In the additive error model \eqref{eq:model-additive} under independent Gaussian errors and one-sided alternatives \eqref{eq:global-test-one-sided},
\cite{arias2017distribution} showed that a phase transition exists for  the approximate support recovery risk \eqref{eq:risk-approximate}.
The \ac{BH} procedure \cite{benjamini1995controlling}, and the Barber-Cand\`es procedure \citep{barber2015controlling} was identified to be asymptotically optimal. % \citet{rabinovich2017optimal} further established the rate-optimality of both procedures under the same regime.
However, \cite{arias2017distribution}, as all related work so far, assumed the non-nulls to follow a common alternative distribution.
We derive a new phase transition result that relaxes this assumption on the alternatives in Chapter \ref{chap:phase-transitions}.

@@

{\bf Asymmetric statistical risks.}
Although weighted sums of false discovery and non-discovery have been studied in the literature mentioned above, the case of simultaneous family-wise error control and marginal, location-wise power requirements, has not been previously considered. 
As a result, asymmetric statistical risks such as \eqref{eq:risk-exact-approx} and \eqref{eq:risk-approx-exact} have not 
been investigated.
As argued in Section \ref{sec:risks}, the properties of these asymmetric risks are of important practical concern in applications such as GWAS.
We study the asymptotic behavior of these risks in Chapters \ref{chap:phase-transitions} and \ref{chap:GWAS} of this text.
% ; the results therein have appeared in \cite{gao2019five}.
% these results have appeared in \cite{gao2018fundamental} and \cite{gao2019five}

@@

{\bf Chi-square models and GWAS.}
The high-dimensional chi-square model \eqref{eq:model-chisq} seemed to have received little attention in the literature.
While the sparse signal detection problem in the chi-square model has been studied \cite{donoho2004higher}, to the best of our knowledge,  asymptotic limits of the support recovery problems have not been studied.
The chi-squared model and the motivating GWAS application are analyzed in Chapter \ref{chap:GWAS}.
The results obtained therein help us explain a phenomenon in GWAS where statistical power decays sharply as function of sample size when the latter is in a small region known as the {\em steep part of the power curve}.  
This empirical fact has long been observed by statistical geneticists but has not been mathematically quantified.  \cite{gao2019upass} provide further details on the power and design in GWAS as well as an accompanying interactive statistical software \citep{gaoUPASS_ShinyApp}.

### 2021-01-13 3.phase_transitions/3.global-tests.tex

The optimality of sparse signal detection was first studied by \citet{ingster1998minimax}, who showed that a phase transition in the $r$-$\beta$ plane exists for the signal detection problem. 
Specifically, assuming equal signal sizes with magnitude $\sqrt{2{r}\log{p}}$, if the signal size $r$ is above a so-called detection boundary,
\begin{equation} \label{eq:detection-boundary-large-signals}
    f(\beta) = 
    \begin{cases}
        \max\{0,\; \beta - 1/2\} & 0 < \beta \le 3/4, \\
        \left(1 - \sqrt{1-\beta}\right)^2 & 3/4 < \beta \le 1,
    \end{cases}
\end{equation} 
then the global null hypothesis $\mu(i)=0$ for all $i=1,\ldots,p$ can be distinguished from the alternative as $p\to\infty$ in the sense of \eqref{eq:support-recovery-success} using the likelihood ratio test; 
otherwise, when signal sizes fall below the boundary, no test can do better than a random guess.
Adaptive tests such as Tukey's \ac{HC} \citep{donoho2004higher} and a modified goodness-of-fit test statistic of \citet{zhang2002powerful} have been identified to attain this performance limit without knowledge of the sparsity and signal sizes. 
% We visualize this boundary $f$ in the upper panel of Figure \ref{fig:phase-diagram-signal-detection}.
It is also known that the max-statistic \eqref{eq:max-statistic} is only efficient when $r>(1+\sqrt{1-\beta})^2$, and is therefore sub-optimal for denser signals where $1/2\le\beta\le 3/4$; see \cite{cai2011optimal}.
In contrast, the sum-of-square-type statistics such as $L_2$ was shown in \cite{fan1996test} to be asymptotically powerless when the $L_2$-norm of the signal $\|\mu\|_2^2$ is $o(\sqrt{p})$, or equivalently, when $\beta>1/2$ in our parametrization.

Notice that the scaling for the signal magnitude $\Delta = \sqrt{2r\log{p}}$ is useful for studying very sparse signals ($\beta>1/2$), but fails to reveal the difficulties of the detection problems when signals are relatively dense ($\beta<1/2$).
Indeed, a different scaling is needed for small but dense signals.
With slight overloading of notation, we parametrize signal sizes as 
\begin{equation} \label{eq:signal-size-small} 
    \underline{\Delta} = p^{\underline{r}}
    \le \mu(i) \le
    \overline{\Delta} = p^{\overline{r}}, \quad \text{for all}\;\;i\in S_p,
\end{equation}
where $\underline{r}$ and $\overline{r}$ are negative constants.
In this scaling for the faint signals, \citet{cai2011optimal} showed that a similar phase transition characterized by the following boundary,
\begin{equation} \label{eq:detection-boundary-small-signals}
    f_s(\beta) = \beta - 1/2, \quad 0 < \beta \le 1/2,
\end{equation} 
exists.
Specifically, if $\overline{r}<f_s(\beta)$, the signal detection fails in the sense of \eqref{eq:support-recovery-failure} regardless of the procedures, while the \ac{HC} statistic continues to attain asymptotically perfect detection when $\underline{r}>f_s(\beta)$. 
We visualize this boundary in the lower panel of Figure \ref{fig:phase-diagram-signal-detection}.

\begin{figure}
      \centering
      \includegraphics[width=0.7\textwidth]{figures/phase_diagram_signal_detection.eps}
      \includegraphics[width=0.7\textwidth]{figures/phase_diagram_signal_detection_vanishing_signals.eps}
      \caption{The phase diagrams of the sparse signal detection problem. 
      Signal size and sparsity are parametrized by $r$ and $\beta$, respectively.
      % Signal sparsity $|S|$ is parametrized by $\beta$. 
      % Signal sizes $\Delta$, parametrized by $r$, increases with dimension $p$ inside the upper panel, and decreases with $p$ inside the power panel.
      The diagrams illustrate the regions where the signal detection problem can be solved asymptotically by some of the commonly used statistics: the maximum ($M$), the sum-of-squares ($L_2$), the sum-of-absolute values ($L_1$), and the sum ($S$).
      In each region of the diagram, the annotated statistics can make the detection risk \eqref{eq:risk-detection} vanish, as dimension $p$ diverges. Conversely, the risks has liminf at least one.
      The detection problem is unsolvable for very sparse and weak signals in the undetectable regions.
      Notice that the $L_1$ and $L_2$ statistics are in fact sub-optimal for all sparsity levels.
      On the other hand, the max-statistic remains powerful for sparse signals ($\beta>1/2$), and is fully efficient when the problem is very sparse ($\beta\ge3/4$).
      The \ac{HC} statistic can detect signals in all configurations in the detectable regions.
      See text and Theorem \ref{thm:detection-optimality}.}
      \label{fig:phase-diagram-signal-detection}
\end{figure}

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and $S$ in this weak signal setting have not been reported in the literature perhaps due to a perceived lack of novelty.
Our first Theorem investigates the performance of these routine statistics for detecting sparse signals in high-dimensions, and summarizes the known results.

@@

Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:proofs} below.
We visualize the results in Theorem in Figure \ref{fig:phase-diagram-signal-detection}.
It is worth noting that the $\beta$-$r$ parameter regions where $L_1$ and $L_2$ statistics are asymptotically powerful coincide, and these statistics are theoretically suboptimal for both sparse regimes ($\beta>1/2$) and relatively dense regimes ($\beta\le 1/2$).

>>>>

The optimality of sparse signal detection was first studied by \citet{ingster1998minimax}, who showed that a phase transition in the $r$-$\beta$ plane exists for the signal detection problem. 
Specifically, consider the so-called {\em detection boundary} function:
\begin{equation} \label{eq:detection-boundary-large-signals}
    f(\beta) = 
    \begin{cases}
        \max\{0,\beta -1/2\} & 0<\beta\le 3/4\\
        \left(1 - \sqrt{1-\beta}\right)^2 & 3/4 < \beta \le 1.
    \end{cases}
    \quad\quad \beta\in (0,1].
\end{equation} 
Assume that the non-zero signal sizes are all equal and parameterized as $\sqrt{2{r}\log{p}}$.
If the signal size parameter $r$ is {\em above} the detection boundary, i.e., $r>f(\beta)$, then the global null 
hypothesis $\mu(i)=0$ for all $i=1,\dots,p$ can be distinguished from the alternative as $p\to\infty$ in the sense of \eqref{eq:support-recovery-success} using the likelihood ratio test.  Otherwise, when the signal sizes fall below the boundary, i.e., $r< f(\beta)$, no test can do better than a 
random guess.  We visualize the detection boundary in the upper panel of Figure
\ref{fig:phase-diagram-signal-detection}.

Adaptive tests such as Tukey's \ac{HC} in \eqref{eq:HC-statistic} \citep{donoho2004higher} and a modified 
goodness-of-fit test statistic of \citet{zhang2002powerful} have been identified to attain this performance limit 
without knowledge of the sparsity and signal sizes. It is also known that the max-statistic \eqref{eq:max-statistic} is only efficient when $r>(1+\sqrt{1-\beta})^2$, and is therefore sub-optimal for denser signals where $1/2\le\beta\le 3/4$; see \cite{cai2011optimal}.
In contrast, the sum-of-square-type statistics such as $L_2$ was shown in \cite{fan1996test} to be asymptotically powerless when the $L_2$-norm of the signal $\|\mu\|_2^2$ is $o(\sqrt{p})$, or equivalently, when $\beta>1/2$ in our parametrization.

Notice that the scaling for the signal magnitude $\Delta = \sqrt{2r\log{p}}$ is useful for studying very sparse signals ($\beta>1/2$), but fails to reveal the difficulties of the detection problems when the signals are relatively dense 
($\beta<1/2$).  This is because $f(\beta) = 0,\ \beta\in (0,1/2]$. Thus, a different scaling is needed to study the regime of
small but dense signals.  In this case, with slight overloading of notation, we parametrize signal sizes as 
\begin{equation} \label{eq:signal-size-small} 
    \underline{\Delta} = p^{\underline{r}}
    \le \mu(i) \le
    \overline{\Delta} = p^{\overline{r}}, \quad \text{for all}\;\;i\in S_p,
\end{equation}
where $\underline{r}$ and $\overline{r}$ are {\rm negative} constants and 
the signal magnitude vanishes, as $p\to\infty$.
In this scaling, for the so-called faint signal regime, \citet{cai2011optimal} established a phase 
transition result characterized by the following boundary,
\begin{equation} \label{eq:detection-boundary-small-signals}
    f_s(\beta) = \beta - 1/2, \quad 0 < \beta \le 1/2.
\end{equation} 
Specifically, if $\overline{r}<f_s(\beta)$, the signal detection fails in the sense of \eqref{eq:support-recovery-failure} regardless of the procedures, while the \ac{HC} statistic continues to attain asymptotically perfect detection when $\underline{r}>f_s(\beta)$. 
We visualize this boundary in the lower panel of Figure \ref{fig:phase-diagram-signal-detection}.

\begin{figure}
      \centering
      \includegraphics[width=0.7\textwidth]{figures/phase_diagram_signal_detection.eps}
      \includegraphics[width=0.7\textwidth]{figures/phase_diagram_signal_detection_vanishing_signals.eps}
      \caption{The phase diagrams of the sparse signal detection problem. 
      Signal size and sparsity are parametrized by $r$ and $\beta$, respectively.
      % Signal sparsity $|S|$ is parametrized by $\beta$. 
      % Signal sizes $\Delta$, parametrized by $r$, increases with dimension $p$ inside the upper panel, and decreases with $p$ inside the power panel.
      The diagrams illustrate the regions where the signal detection problem can be solved asymptotically by some of the commonly used statistics: the maximum ($M$), the sum-of-squares ($L_2$), the sum-of-absolute values ($L_1$), and the sum ($T$).
      In each region of the diagram, the annotated statistics can make the detection risk \eqref{eq:risk-detection} vanish, as dimension $p$ diverges. Conversely, the risks has liminf at least one.
      The detection problem is unsolvable for very sparse and weak signals in the undetectable regions.
      Notice that the $L_1$ and $L_2$ statistics are in fact sub-optimal for all sparsity levels.
      On the other hand, the max-statistic remains powerful for sparse signals ($\beta>1/2$), and is fully efficient when the problem is very sparse ($\beta\ge3/4$).
      The \ac{HC} statistic can detect signals in all configurations in the detectable regions.
      See text and Theorem \ref{thm:detection-optimality}.}
      \label{fig:phase-diagram-signal-detection}
\end{figure}

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and 
\stilian{the sum statistic $T$ in \eqref{eq:sum-statistic}}{R2, Typo 7}
in this weak signal setting have not been reported in the literature. %perhaps due to a perceived lack of novelty. 
Our first theorem establishes the performance of these simple but popular statistics for detecting sparse signals 
in high-dimensions, and summarizes the known results.

@@

Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:supplement}.
We visualize the results in Theorem in Figure \ref{fig:phase-diagram-signal-detection}.
It is worth noting that the $\beta$-$r$ parameter regions where $L_1$ and $L_2$ statistics are asymptotically powerful coincide, and these statistics are theoretically suboptimal for both sparse regimes ($\beta>1/2$) and relatively dense regimes ($\beta\le 1/2$).

### 2021-01-13 3.phase_transitions/3.proofs-phase-transitions.tex

\subsection{Auxiliary facts of Gaussian distributions}
\label{sec:Gaussian-distributions}
\input{3.phase_transitions/3.Gaussian-distributions.tex}

\subsection{Monotonicity of the Benjamini-Hochberg procedure}
\label{suppsec:BH-monotonicity}
\input{3.phase_transitions/3.BH_monotonicity.tex}

\subsection{Proof of Theorem \ref{thm:detection-optimality}}
\label{subsec:proof-additive-error-detection-boundaries}

@@

\subsection{Proof of Theorem \ref{thm:Gaussian-error-approx-boundary}}
\label{subsec:proof-additive-error-approx-boundaries}

@@

\subsection{Proof of Theorems \ref{thm:Gaussian-error-exact-approx-boundary} and \ref{thm:Gaussian-error-approx-exact-boundary}}
\label{subsec:proof-additive-error-mix-boundaries}

>>>>

\section{Auxiliary facts of Gaussian distributions}
\label{sec:Gaussian-distributions}
\input{3.phase_transitions/3.Gaussian-distributions.tex}

\section{Monotonicity of the Benjamini-Hochberg procedure}
\label{suppsec:BH-monotonicity}
\input{3.phase_transitions/3.BH_monotonicity.tex}

\section{Proof of Theorem \ref{thm:detection-optimality}}
\label{subsec:proof-additive-error-detection-boundaries}

@@

\section{Proof of Theorem \ref{thm:Gaussian-error-approx-boundary}}
\label{subsec:proof-additive-error-approx-boundaries}

@@

\section{Proof of Theorems \ref{thm:Gaussian-error-exact-approx-boundary} and \ref{thm:Gaussian-error-approx-exact-boundary}}
\label{subsec:proof-additive-error-mix-boundaries}
